COFFEE: an Optimizing Compiler\\for Finite Element Local Assembly

### 2014-04-03 paper/bare_conf.tex
Some rewording of intro section

\section{Introduction}
In many fields, like computational fluid dynamics, computational electromagnetics, and structural mechanics, phenomena are modelled by means of partial differential equations (PDEs). Numerical techniques, like finite volume method and finite element method, are widely-employed to approximate solutions of PDEs. Unstructured meshes are often used to discretize the equation domain, since their geometric flexibility allows solvers to be extremely effective. The solution is sought in each element of the discretized domain by applying suitable numerical kernels. As the number of elements can be of the order of millions, a major issue is the time required to execute the computation, which can be hours or days. To address this problem, domain-specific languages (DSLs) have been developed. The successful porting of Hydra, a computational fluid dynamics industrial application devised by Rolls Royce for turbomachinery design (based on Finite Volume Method, roughly 50000 lines of code and mesh sizes that can be over 100 millions edges), to OP2~\cite{pyop2isc}, demonstrates the effectiveness of the DSL approach for implementing PDEs solvers~\cite{IstvanHydra}.

OP2 adopts a kernel-oriented programming model, in which the computation semantics is expressed through self-contained functions, or ``kernels''. A kernel is applied to all elements in a set of mesh components (e.g. edges, vertices, cells/elements), with an implicit synchronization between the application of two consecutive kernels. On commodity multi-cores, a kernel is executed sequentially by a thread, while parallelism is achieved partitioning the mesh and assigning each partition to a thread. Similar programming and execution models are adopted in~\cite{Fenics},~\cite{fluidity_manual_v4},~\cite{lizst}. Kernel optimization is one of the major concerns in unstructured mesh applications. In this paper, we tackle this problem by proposing an optimization strategy for a class of kernels used in finite element methods.

We focus on local assembly (``assembly'', in the following), a fundamental step of a finite element method that covers an important fraction of the overall computation run-time, often in the range 30$\%$-60$\%$. During the assembly phase, the solution of the PDE is approximated by executing a suitable kernel over all elements in the discretized domain. A kernel's working set is usually small enough to fit the L1 cache; it might need L2 cache when high-order methods are employed to improve the accuracy of the solution. However, we do not consider the latter case. An assembly kernel is characterized by the presence of an affine, often non-perfect loop nest, where individual loops are rather small: their trip count rarely exceeds 30, whereas the minimum value is 3. In the innermost loop, a problem-specific expression evaluates a bi-dimensional array, representing the result of local assembly in an element of the discretized domain. With such a kernel structure, we focus on aspects like minimization of floating-point operations, register allocation and instruction-level parallelism, especially in the form of SIMD vectorization.

@@

Due to the constraints of available compilers and on linear-algebra specialized libraries, we have automated a set of generic and model-driven code transformations in COFFEE\footnote{COFFEE stands for COmpiler For FinitE Element local assembly.}, a compiler for optimizing local assembly kernels. COFFEE is integrated with Firedrake, a system for solving PDEs through the finite element method based on the OP2 abstraction~\cite{firedrake-code}. It supports any problems expressible with this framework. This allows us to evaluate our code transformations in a range of real-world problems, varying key parameters that impact both solution accuracy and kernel cost, namely the polynomial order of the method (from $p=1$ to $p=4$) and the geometry of elements in the discretized domain (2D triangle, 3D tetrahedron, 3D prism).

Early experiments showed that Firedrake-generated code for non-trivial assembly kernels was sub-optimal. Our cost-model-driven sequence of source-to-source code transformations, aimed at improving SIMD vectorization and register data locality, can result in performance improvements up to 1.5$\times$ over ``softly-optimized'' code (i.e. where only basic transformations are performed, such as generalized loop-invariant code motion, padding, and data alignment), and up to 4.44$\times$ over original kernels. The contributions of this paper are threefold
\begin{itemize}
\item An optimisation strategy for a class of kernels widely-used in scientific applications, namely local assembly in the context of the finite element method. Our approach, which also exploits domain knowledge, goes beyond the limits of both vendor and research compilers.
\item Design and implementation of a compiler that automates the proposed code transformations for any problems expressible in Firedrake.
\item Systematic evaluation using a suite of examples of real-world importance, and evidence of significant performance improvements on two Intel machines, a Sandy Bridge and the Phi.
\end{itemize}

The paper is organized as follows. In Section~\ref{sec:background} we provide some background on local assembly, showing code generated by Firedrake and emphasizing critical computational aspects. Section~\ref{sec:code-transf} describes in detail the various code transformations, highlighting when and how domain-knowledge has been exploited. The design and implementation of our compiler is discussed in Section~\ref{sec:pyop2-compiler}. Section~\ref{sec:perf-results} shows performance results. Related work are illustrated in Section~\ref{sec:related-work}, while Section~\ref{sec:conclusions} concludes the paper.

>>>>

\section{Introduction}
In many fields, such as computational fluid dynamics, computational electromagnetics, and structural mechanics, phenomena are modeled by means of partial differential equations (PDEs). Numerical techniques, such as the finite volume and finite element method, are widely employed to approximate solutions of PDEs. Unstructured meshes are often used to discretize the computational domain, since their geometric flexibility allows adapting the mesh to the required resolution. The solution is sought in each element of the discretized domain by applying suitable numerical kernels. As the number of elements can be of the order of millions, the time required to execute the computation can be hours or days. To address this problem, domain-specific languages (DSLs) have been developed. The successful porting of Hydra, a computational fluid dynamics industrial application devised by Rolls Royce for turbomachinery design to OP2~\cite{pyop2isc} demonstrates the effectiveness of the DSL approach for implementing PDEs solvers~\cite{IstvanHydra}. Hydra uses the finite volume method, consists of roughly 50000 lines of code and is able to handle mesh sizes of over 100 millions edges.

% FIXME: that citation is wrong! You're talking about OP2. You MUST maintain that distinction

OP2 adopts a programming model, in which computations are expressed through self-contained functions, referred to as ``kernels''. A kernel is applied to all elements in a set of mesh entities such as edges, vertices, or cells, with an implicit synchronization between the application of two consecutive kernels. On commodity multi-cores, a kernel is executed sequentially by a thread, while parallelism is achieved partitioning the mesh and assigning each partition to a thread. Similar programming and execution models are adopted in~\cite{Fenics},~\cite{fluidity_manual_v4},~\cite{lizst}. Kernel optimization is one of the major concerns in unstructured mesh applications. In this paper, we tackle this problem by proposing an optimization strategy for a class of kernels used in finite element methods.

% FIXME: citation for the 30-60% figure?

We focus on local assembly (``assembly'' in the following), a fundamental step of the finite element method that covers an important fraction of the overall computation time, often in the range of 30$\%$-60$\%$. During the assembly phase, the solution of the PDE is approximated by executing a suitable kernel over all elements in the discretized domain. A kernel's working set is usually small enough to fit the L1 cache; it might need L2 cache when high-order methods are employed to improve the accuracy of the solution. However, we do not consider the latter case. An assembly kernel is characterized by the presence of an affine, often non-perfect loop nest, where individual loops are rather small: their trip count rarely exceeds 30, and may be as low as 3 for low-order methods. In the innermost loop, a problem-specific expression evaluates a two dimensional array, representing the result of local assembly in an element of the discretized domain. With such a kernel structure, we focus on aspects such as minimization of floating-point operations, register allocation and instruction-level parallelism, especially in the form of SIMD vectorization.

@@

Due to the constraints of available compilers and specialized linear-algebra libraries, we have automated a set of generic and model-driven code transformations in the COmpiler For FinitE Element local assembly (COFFEE), a compiler for optimizing local assembly kernels. COFFEE is integrated with Firedrake~\cite{firedrake-code}, a system for solving PDEs through the finite element method based on the PyOP2 abstraction~\cite{pyop2isc,pyop2ws}. It supports any problems expressible with this framework. This allows us to evaluate our code transformations in a range of real-world problems, varying key parameters that impact both solution accuracy and kernel cost, namely the polynomial order of the method (from $p=1$ to $p=4$) and the geometry of elements in the discretized domain (2D triangle, 3D tetrahedron, 3D prism).

% FIXME: You're talking about FFC-generated code really

Early experiments showed that Firedrake-generated code for non-trivial assembly kernels was sub-optimal. Our cost-model-driven sequence of source-to-source code transformations, aimed at improving SIMD vectorization and register data locality, can result in performance improvements up to 1.5$\times$ over ``softly-optimized'' code (that is code where only basic transformations are performed, such as generalized loop-invariant code motion, padding, and data alignment), and up to 4.44$\times$ over original kernels. The contributions of this paper are threefold:
\begin{itemize}
\item An optimization strategy for finite element local assembly, a class of kernels widely-used in scientific applications. Our approach exploits domain knowledge and goes beyond the limits of both vendor and research compilers.
\item Design and implementation of a compiler that automates the proposed code transformations for any problems expressible in Firedrake.
\item Systematic evaluation using a suite of examples of real-world importance, and evidence of significant performance improvements on two Intel architectures, a Sandy Bridge CPU and the Xeon Phi.
\end{itemize}

This paper is organized as follows: In Section~\ref{sec:background} we provide some background on local assembly, showing code generated by Firedrake and emphasizing important computational aspects. Section~\ref{sec:code-transf} describes the various code transformations in detail, highlighting when and how domain knowledge has been exploited. The design and implementation of our compiler is discussed in Section~\ref{sec:pyop2-compiler}. Section~\ref{sec:perf-results} shows performance results. Related work are illustrated in Section~\ref{sec:related-work}, while Section~\ref{sec:conclusions} concludes the paper.
